---
title: 'Figures for Essential Measures: Ancestry, Race, and Social Difference'
author: "Aaron Gullickson"
output: 
  pdf_document: 
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#read in necessary files
anc.single <- read.csv("output/entropy_single_2000.csv")
anc.big <- read.csv("output/entropy_omb_2000.csv")
anc.big.byyear <- read.csv("output/entropy_omb_byyear.csv")
```

##Introduction

This program will reproduce the figures used in the article "[Essential Measures: Ancestry, Race, and Social Difference](http://abs.sagepub.com/content/60/4/498)" published in the American Behavioral Scientist.

The production of Figures 1-3 is dynamic and will input the currently processed CSV files for entropy, but Figures 4 and 5 will have to have the values manually changed based on the results from the models if changes are made. Those models are not read in directly because they are calculated in Stata. 

##Figures

```{r echo=FALSE, fig.cap="Race-reporting inconsistency time trend by big racial ancestry, 1990 to 2012"}
multi <- anc.big.byyear$rancestry!="White" & anc.big.byyear$rancestry!="Black" & 
  anc.big.byyear$rancestry!="Indig" & anc.big.byyear$rancestry!="Asian" & anc.big.byyear$rancestry!="Latino"

multi.weighted.average <- tapply(anc.big.byyear$entropy[multi]*anc.big.byyear$N[multi],anc.big.byyear$year[multi],sum)/tapply(anc.big.byyear$N[multi],anc.big.byyear$year[multi],sum)

table.big <- tapply(anc.big.byyear$entropy, anc.big.byyear[,1:2], sum, na.rm=T)[c("White","Black","Indig","Asian","Latino"),]
table.big <- rbind(table.big, multi.weighted.average)
rownames(table.big)[[6]] <- "Multiple"
rownames(table.big)[[3]] <- "Indigenous"

year <- as.numeric(colnames(table.big))
par(mar=c(5,4,2,2))
plot(year, table.big[1,], lwd=6, lty=1, xlab="year",ylab="race reporting inconsistency", ylim=c(0,.85), type="l", las=1, xlim=c(1987,2012), xaxt="n")
axis(1, at=c(1990, 2000, 2010))
lines(year, table.big[2,], lwd=3, lty=2)
lines(year, table.big[3,], lwd=3, lty=3)
lines(year, table.big[4,], lwd=3, lty=4)
lines(year, table.big[5,], lwd=3, lty=5)
lines(year, table.big[6,], lwd=3, lty=1, col="grey")
text(1990,table.big[,1],rownames(table.big), cex=0.7, pos=2)
```

```{r echo=FALSE, fig.cap="Race-reporting inconsistency by big race category among single and multiple racial ancestry respondents, 2000 to 2012"}
anc.big$colors <- c(rep("grey20",5),rep("grey70",10))
anc.big$names <- c("White","Black","Indigenous", "Asian", "Latino", "White/Black","White/Indigenous", "White/Asian", "White/Latino", "Black/Indigenous", "Black/Asian", "Black/Latino", "Indigenous/Asian", "Indigenous/Latino", "Asian/Latino")
anc.big <- anc.big[order(anc.big$entropy, decreasing=FALSE),]
par(mar=c(5,8,1,1))
barplot(anc.big$entropy, names.arg=anc.big$names, las=1, xlim=c(0,0.8), horiz=TRUE, col=anc.big$colors, xlab="race reporting inconsistency")
legend(0.4, 4, legend=c("Single racial ancestry","Multiple racial ancestry"), fill=c("grey20","grey70"),cex=0.85)
```

```{r echo=FALSE, fig.cap="Boxplot of race-reporting inconsistency by individual ancestry among single ancestry respondents clustered by meso racial ancestry, 2000 to 2012"}
#FIGURE 3 boxplots
#ancestry names
rancestry.names=c("European","African\nAmerican","American\nIndian  ","E and\nSE Asian","Pacific\nIslander","Latino","Afro-\nCaribbean","M East/\nN African","African","South\nAsian")
#color by OMB category
rancestry.colors <- c("white","grey20","grey60","grey80","grey60","grey40","grey20","white","grey20","grey80")
names(rancestry.colors) <- rancestry.names
anc.single$rancestry1names <- as.factor(rancestry.names[anc.single$meso])
anc.single$rancestry1.ordered <- with(anc.single, reorder(as.factor(rancestry1names), entropy, median))
par(mar=c(4,4,1,1))
bx <- boxplot(anc.single$entropy~anc.single$rancestry1.ordered, las=1,
              col=rancestry.colors[levels(anc.single$rancestry1.ordered)], 
              pch=20, ylim=c(0,0.8), 
              ylab="race reporting inconsistency", cex.axis=0.6, cex.lab=0.7,
              xlab="Meso racial ancestry")
#report outliers individually
text(bx$group+0.1, bx$out, c("Icelander","Portugese","German from Russia","Roma","Uzbek","New Zealander","Cape Verdean","Sudanese","Creole","South American","Moroccan","Other Arab","Eurasian","Amerasian","Afghan"), pos=4, cex=0.5, offset=0.01)
legend(1,0.8,legend=c("White","Black","Indigenous","Asian","Latino"), fill=c("white","grey20","grey60","grey80","grey40"), cex=0.35, title="Big race categories")
```

```{r echo=FALSE, fig.cap="Partitioning of the overall ancestry inequality in college completion from 2000 to 2012 by big race, meso race, and individual ancestry components"}
par(mfrow=c(2,1), mar=c(3,4,1,1))
prop <- c(71.4, 3.3, 25.3)
bp<- barplot(prop, beside=TRUE, names.arg=c("Big Race\n ", "Meso Race\nwithin Big Race", "Individual Ancestry\nwithin Meso Race"),
        ylim=c(0,90), main="Panel A: Share of Overall Ancestry Inequality")
height <- prop/2
height[2] <- -1#height[2]/15
text(bp, prop, paste(prop,"%", sep=""), pos=3)

prop2 <- c(58.4, 2.8, 0.4, 15.5, 23.1)
names(prop2) <- c("White","Black","Indigenous","Asian","Latino")
prop2 <- sort(prop2, decreasing=TRUE)
bp2<- barplot(prop2, ylim=c(0,90), main="Panel B: Share of Individual Ancestry Inequality by Big Race")
text(bp2, prop2, paste(prop2,"%", sep=""), pos=3)
```

```{r echo=FALSE, fig.cap="Average big race differences in college completion and within big race variation by individual ancestry, 2000 to 2012"}
par(mfrow=c(1,1), mar=c(3,4,1,1), las=1, bty="n")
baselodds <- c(-.287, -1.774, -2.448, .069, -1.403)
se <- c(0.532,0.461, 0.552, 0.784, 0.804)
baseodds <- exp(baselodds)
upper <- exp(baselodds+1.96*se)
lower <- exp(baselodds-1.96*se)
plot(1:5, baseodds, pch=21, log="y", ylim=c(min(lower),max(upper)), xaxt="n", xlim=c(0.5,5.5), cex=2, bg="grey", ylab="Odds of college completion")
epsilon <- 0.05
segments(1:5, lower, 1:5, upper, lwd=2)
segments(1:5-epsilon, lower, 1:5+epsilon, lower, lwd=2)
segments(1:5-epsilon, upper, 1:5+epsilon, upper, lwd=2)

axis(1, at=1:5, labels=c("White","Black","Indigenous","Asian","Latino"))
```